
---
title: "PA1_template.Rmd"
author: "Roshan"
date: "December 30, 2018"
output: 
  html_document:
    keep_md: TRUE
---

```{r setoptions, echo = FALSE}

library(knitr)
opts_chunk$set(echo = FALSE, results = "hide")

```

```{r load libraries, message=FALSE, warning=FALSE, paged.print=FALSE,echo=TRUE}

#setwd("./Project")
library(data.table)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)

```
```{r process data, echo = TRUE}

#read data from the csv file
originalData = read.csv("activity.csv")

#change the date as Date class and format "%m-%d-%Y"
originalData$date <- as.Date(originalData$date,format = "%m/%d/%Y")
originalData$date <- format.Date(originalData$date,"%m-%d-%Y")

#remove NA rows from the data
originalData<-originalData[!is.na(originalData$steps),]

```

###[1] What is mean total number of steps taken per day?

```{r Plot the histogram of the total number of steps taken each day, echo=TRUE}

#read the original data from csv file
 originalData = read.csv("activity.csv")
 
#change the date as Date class and format "%m-%d-%Y"
 originalData$date <- as.Date(originalData$date,format = "%m/%d/%Y")
originalData$date <- format.Date(originalData$date,"%m-%d-%Y")

#select only !NA values
originalData<-originalData[!is.na(originalData$steps),]

#keep seperate calculation for for with NA
originalDataNA = read.csv("activity.csv")

# calculate sum of steps per date
data <- originalData %>%
        group_by(date) %>%
        summarise(totalSteps = sum(steps))%>%
        ungroup()
#construct the plot 
g <- ggplot(data, aes(data$totalSteps)) + 
  geom_histogram(breaks=seq(10, 22000, by=1000),col="grey",fill= "red") +
   ggtitle("Total Steps Counts per Date") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("Interval") + 
  ylab("Total Step Counts")

#calculate the mean and mean of total steps
TotStepsMedian <- as.integer(median(data$totalSteps))
TotStepsMean <- as.integer(mean(data$totalSteps))

```


#### histogram of the total number of steps taken each day

```{r plot}
g
```


  Mean total-number of steps taken per day: **`r TotStepsMedian`**

  Median total number of steps taken per day: **`r TotStepsMean`**


###[2] What is the average daily activity pattern?

```{r Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days , echo = TRUE}

#calculate average of steps
aveDailyActivity <- originalData %>% 
                    group_by(interval) %>% 
                    summarise(aveSteps = mean(steps))

#calculate maximum of average steps
maxX <- max(aveDailyActivity$aveSteps)

#calcuate average daily activities with maximum aveSteps
maxY <- aveDailyActivity[aveDailyActivity$aveSteps==maxX,]
 
#set label
labels  = paste("[", paste(as.numeric(maxY[1,1]),
#set all maxY as numeric                           
ceiling(as.numeric(maxY[1,2])),sep=","),"]")

#plot interval vs aveSteps
g2 <- ggplot(aveDailyActivity, aes(aveDailyActivity$interval,aveDailyActivity$aveSteps )) + geom_line() +
  ggtitle("Average Steps Over 5min Intervals") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("Time Interval (5min unites)") + ylab("Average Steps") +
  geom_text(aes(as.numeric(maxY[1,1] + 300), as.numeric(maxY[1,2])),label= labels)

```
####The average daily activity pattern

```{r the average daily activity pattern}
g2

```

####Maximum number of steps - on average across all the days

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 
                          **`r as.numeric(maxY[1,1])`**
                          
###[3]Imputing missing values

####Calculate and report the total number of missing values in the dataset 

```{r Calculate and report the total number of missing values,echo = TRUE}

#count total NA value rows in original data with NA
countsNA = count(originalDataNA,is.na(originalDataNA$step))

#set count numbers as numeric
totalNA <- as.numeric(countsNA[2,2])

```

The total number of rows with NAs is **`r totalNA`**


####Devise a strategy for filling in all of the missing values in the dataset

```{r Devise a strategy for filling in all of the missing values in the dataset, echo = TRUE}

#read dataset again to make sure itis original
originalData = read.csv("activity.csv")

#change the date as Date class and format "%m-%d-%Y"
originalData$date <- as.Date(originalData$date,format = "%m/%d/%Y")
originalData$date <- format.Date(originalData$date,"%m-%d-%Y")

#original Data without NAs
originalDataWithoutNA<-originalData[!is.na(originalData$steps),]

#read data to use with NAs
dataWithNA = read.csv("activity.csv")

#change the date as Date class and format "%m-%d-%Y"
dataWithNA$date <- as.Date(dataWithNA$date,format = "%m/%d/%Y")
dataWithNA$date <- format.Date(dataWithNA$date,"%m-%d-%Y")

#calculate the average steps without NAs over interval
aveDailyActivityWtNA <- originalDataWithoutNA %>% 
                      dplyr::group_by(interval) %>% 
                      dplyr::summarise(aveSteps = mean(steps))

aveDailyActivityWtNA$aveSteps <- ceiling(aveDailyActivityWtNA$aveSteps )

#sum over new aveDailyActivityWtNA set aveSteps
dataNA<- aveDailyActivityWtNA %>%
    dplyr::summarise(totalSteps = sum(aveSteps))

dataNA$totalSteps <- ceiling(dataNA$totalSteps)

#calculate sum over all original data with NA
 data <- dataWithNA %>%
   group_by(date) %>%
   summarise(totalSteps = sum(steps))%>%
   ungroup()
 
 #mutate dataset that is equal to the original dataset but with the missing data
 data <- data %>% mutate(totalSteps = ifelse(is.na(totalSteps), dataNA, totalSteps))
 
 #set totalSteps as integer to match origina type -- NEW Data
 data$totalSteps <- as.integer(data$totalSteps)

  #construct the plot
g3 <- ggplot(data, aes(data$totalSteps)) + 
      geom_histogram(breaks=seq(10, 22000, by=1000),col="grey",fill= "red") +
      labs(title = "                      Total Steps Counts per Date") +
      xlab("Total Steps") + 
      ylab("Total Step Counts")
```


#### Histogram of the total number of steps taken each day with NA replaced

```{r total number of steps taken each day with NA replaced}
g3
```

#### Mean and Median total number of steps taken per day with replaced NAs

```{r }
TotStepsMedianNA <- as.integer(median(data$totalSteps))
TotStepsMeanNA <- as.integer(mean(data$totalSteps))

```
 Mean total-number of steps taken per day with replaced NAs: **`r TotStepsMedianNA`**

 Median total number of steps taken per day with replaced NAs: **`r TotStepsMeanNA`**

#### Difference in Mean and Median total number of steps taken per day w/out replaced NAs
 Mean difference = **`r sum(TotStepsMedianNA - TotStepsMedian)`**

 Median difference = **`r sum(TotStepsMeanNA - TotStepsMean)`**

 Percentage of mean diffrence =  **`r (sum(TotStepsMedianNA - TotStepsMedian)/TotStepsMedian)*100`**

 Percentage of median diffrence =  **`r (sum(TotStepsMeanNA - TotStepsMean)/TotStepsMean)*100`**
 
 ! Replacing NAs with averaged sum over interals wont impact significatly on data


###Are there differences in activity patterns between weekdays and weekends?

```{r  differences in activity patterns between weekdays and weekends}

#again read the original data make sure they are original from previous calculations
originalData = read.csv("activity.csv")

#change the date as Date class and format "%m-%d-%Y"
originalData$date <- as.Date(originalData$date,format = "%m/%d/%Y")
originalData$date <- format.Date(originalData$date,"%m-%d-%Y")

#original data without NAs
originalDataWithoutNA<-originalData[!is.na(originalData$steps),]

#for data  calculations with NAs
dataWithNA = read.csv("activity.csv")

#change the date as Date class and format "%m-%d-%Y"
dataWithNA$date <- as.Date(dataWithNA$date,format = "%m/%d/%Y")
dataWithNA$date <- format.Date(dataWithNA$date,"%m-%d-%Y")

#mean average over intervals without NAs
aveDailyActivityWtNA <- originalDataWithoutNA %>% 
  dplyr::group_by(interval) %>% 
  dplyr::summarise(aveSteps = mean(steps))

aveDailyActivityWtNA$aveSteps <- ceiling(aveDailyActivityWtNA$aveSteps )

#filter NAs from original dta
filt<- filter(dataWithNA,is.na(dataWithNA$steps))

#fill average activity steps into filtered NAs 
filt$steps <- aveDailyActivityWtNA$aveSteps


#bind both originalDataWithoutNA with filterd and replaced NAs
finalData <- rbind(originalDataWithoutNA,filt)
#arrange the data according to date order
finalData<-arrange(finalData,finalData$date)

#summarise mean steps on finalData
aveDailyActivityWithNA <- finalData %>% 
  dplyr::group_by(interval) %>% 
  dplyr::summarise(aveSteps = ceiling(mean(steps)))

#parse finalData$data as Date format
finalData$date<-as.Date(parse_date_time(finalData$date, c('mdy')))

#label weekdays
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')

#factor weekday and weekend usng weekdays() function
days <- factor((weekdays(finalData$date) %in% weekdays1), 
               levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

#define days data set as data.frame
dataDays<-data.frame(finalData,days)

#summaraise mean over days and interval groups
df <- dataDays %>%  group_by(days,interval) %>% 
  summarise(mean = mean(steps))

#construct the plot
g4 <- ggplot(df, aes(x=interval, y=mean)) +   geom_line() + 
  theme_bw() +
  xlab("Interval") + ylab("Number of steps") + 
  facet_wrap(days~.,ncol =1)
```

####Plot of average number of steps taken, averaged across all weekday days or weekend days

```{r plot average number of steps taken, averaged across all weekday days or weekend days}
g4

```
